Sample Clustering

### 2017-02-15 Clustering.Rmd
Updated for new patient data
Updating the segmentation script

sum.patient.data = ddply(patient.info, .(Patient, Status), summarise, years=paste(range(Endoscopy.Year), collapse='-'), 
      n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T))

load('patients.Rdata', verbose=T)
```

@@

  x1 = patient.data[[pt]]$seg.vals[, patient.data[[pt]]$info$Samplename]
  x1 = x1[,patient.data[[pt]]$info$Samplename]
  colnames(x1) = paste(patient.data[[pt]]$info$Endoscopy.Year, '(', patient.data[[pt]]$info$Pathology, ')', sep='' )

@@

    if (length(table(mod1dr$mixcomp)) > 1) {
      plot(mod1dr, what = "evalues")
      title(main=pt)
      plot(mod1dr, what = "density")
    }
  })
}

@@

## Using the segmentation values
sim = sum.patient.data[,c(2:4)]
rownames(sim) = sum.patient.data[,1]
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
    plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
    plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))
  #heatmap(ac)
}
sim$id = rownames(sim)

@@

#t.test( subset(sim, Status == 'NP')$clusters, subset(sim, Status == 'P')$clusters )$p.value

@@

```{r xval_pt, echo=T, warning=F, message=T, fig.align='left', fig.height=4, eval=T}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

ggplot(melt(data.frame(loo_xval[,c(1,3)])), aes(x=variable, y=value, fill=variable)) +  
    geom_boxplot() + geom_jitter(shape=16, width=0.2) + geom_hline(yintercept=0.05, color='red', alpha=0.5) +
    stat_boxplot(geom ='errorbar') + labs(y='p-values', x='', title='Leave one (patient) out clustering')
```

@@

```{r xval_smp, echo=F, warning=F, message=T, fig.align='left', fig.height=6, eval=T}

# Leave out samples?
looS_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:20) { 
  simX = sum.patient.data[,c(2:4)]
  rownames(simX) = sum.patient.data[,1]

@@

ggplot(melt(data.frame(looS_xval[,c(2,3)])), aes(x=variable, y=value, fill=variable)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
  geom_hline(yintercept=0.05, color='red', alpha=0.5) +
    stat_boxplot(geom ='errorbar') + labs(y='p-values', x='', title='Leave one (sample) out clustering')
```

@@

```{r nofinal, echo=F, message=F, warning=F, fig.height=6, fig.width=10}
simX = sum.patient.data[,c(2:4)]
rownames(simX) = sum.patient.data[,1]
# Leave out final sample on the progressors

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

ttps = t.test( subset(simX, Status == 'NP')$ss, subset(simX, Status == 'P')$ss )$p.value
gg1 = ggplot(melt(data.frame(simX[c('Status','ss')])), aes(x=Status, y=value, fill=Status)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
    stat_boxplot(geom ='errorbar') + labs(y='sim values', x='', title=paste('Leave out HGD sample(s) p-value', signif(ttps, 3)))

ttpc = t.test( subset(simX, Status == 'NP')$clusters, subset(simX, Status == 'P')$clusters )$p.value
gg2 = ggplot(melt(data.frame(simX[c('Status','clusters')])), aes(x=Status, y=value, fill=Status)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
    stat_boxplot(geom ='errorbar') + labs(y='num. clusters', x='', title=paste('Leave out HGD sample(s) p-value', signif(ttpc, 3)))

grid.arrange(gg1, gg2, ncol=2)

@@

pander(subset(simX, Status == 'P' & ss >= mean(ss)), caption="These patients are more similar to Non-Progressors without their HGD sample(s)")

simX = simX[, c('Status', 'clusters', 'ns','ss')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

@@

pvals = as.data.frame(matrix(ncol=2,nrow=0,dimnames=list(c(), c('clusters','ss'))))
for (qv in seq(0.1,0.9,0.1)) {
  print(qv)

@@

    simPC[pt, 'ns'] = net.similarity(ac)
    simPC[pt, 'ss'] = sum.similarity(ac)
    simPC[pt, 'clusters'] = length(clusters(ac))
  }

  # May be a difference
  ttps = t.test( subset(simPC, Status == 'NP')$ss, subset(simPC, Status == 'P')$ss )$p.value
  ttpn = t.test( subset(simPC, Status == 'NP')$ns, subset(simPC, Status == 'P')$ns )$p.value
  ttpc = t.test( subset(simPC, Status == 'NP')$clusters, subset(simPC, Status == 'P')$clusters )$p.value

  pvals = rbind(pvals,cbind(ttpc,ttps))
}
colnames(pvals) = c('clusters','sum.similarity')
pvals$qv=seq(0.1,0.9,0.1)

ggplot(melt(pvals, id.vars='qv'),aes(x=qv,y=value, color=variable)) + geom_point() +
  labs(main="qvalus for AP", y='pvalue NP vs P', x='q param')

@@

```{r precrec, echo=F, warning=F, message=F, fig.height=12, fig.width=20}

>>>>

load('patients.Rdata', verbose=T)

sum.patient.data = as.data.frame(patient.info %>%
  group_by(Patient, Status) %>%
  summarise(
    years=paste(range(Endoscopy.Year), collapse='-'),
    n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T),
    mean.reads=mean(Total.Reads),
    sd.reads=sd(Total.Reads)
  ))

pander(sum.patient.data)
```

# Global View and Questions

```{r global_pt, echo=F, message=F, warning=F}
be = subset(patient.info, Pathology == 'BE')
lgd = subset(patient.info, Pathology <= 'LGD')

p53Status = data.frame(rbind('Non-Progressors'=table(subset(patient.info, Status == 'NP')$p53.Status), 
           'Progressors'=table(subset(patient.info, Status == 'P')$p53.Status)))
colnames(p53Status) = c('None', 'Positive')

ttr = t.test(subset(patient.info, Status == 'NP')$Total.Reads, subset(patient.info, Status == 'P')$Total.Reads)
ttrbe = t.test(subset(be, Status == 'NP')$Total.Reads, subset(be, Status == 'P')$Total.Reads)
ttrlgd = t.test(subset(lgd, Status == 'NP')$Total.Reads, subset(lgd, Status == 'P')$Total.Reads)

tbclgd = t.test(subset(lgd, Status == 'NP')$Barretts.Cellularity, subset(lgd, Status == 'P')$Barretts.Cellularity)
tbcbe = t.test(subset(be, Status == 'NP')$Barretts.Cellularity, subset(be, Status == 'P')$Barretts.Cellularity)

t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, 
               subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity)

t.test(subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity, 
               subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)

## No differences between the number of reads by pathology
t.test(subset(patient.info, Pathology == 'BE')$Total.Reads, subset(patient.info, Pathology == 'LGD')$Total.Reads)
t.test(subset(patient.info, Pathology == 'BE')$Total.Reads, subset(patient.info, Pathology == 'HGD')$Total.Reads)
t.test(subset(patient.info, Pathology == 'LGD')$Total.Reads, subset(patient.info, Pathology == 'HGD')$Total.Reads)

t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity)
t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)
t.test(subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity, subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)

There are a few things we need to be careful of and a couple of questions to check here:

1. The total number of reads per sample between NP and P might show a difference, but currently with multi-test correction it's not significant.
`r  signif(ttr$p.value,2)*11`
So are those samples with lower numbers of reads showing less complexity overall?  When sequencing NP are we likely to get fewer reads than with P? If we look only at the samples with BE (no LGD or HGD samples) the difference disappears `r signif(ttrbe$p.value, 2)*11`. If we include LGD with the BE samples we again see the difference.  `r signif(ttrlgd$p.value, 2)*11`.  This is suggestive that changes in copy number result in more reads being output.

2. There is a significant difference in the cellularity of the Barretts sample, with progressors having a higher mean cellularity, even when only comparing samples that were either LGD or BE.  `r signif(tbclgd$p.value, 2)` However, even the BE only samples show a higher cellularity overall `r signif(tbcbe$p.value, 2)*11`  That said, currently we have a high rate of missing information, with about `r length(which(is.na(subset(patient.info, Status == 'P')$Barretts.Cellularity)))/length(which(patient.info$Status == 'P'))*100`% of the cellualarities missing for progressors and `r length(which(is.na(subset(patient.info, Status == 'NP')$Barretts.Cellularity)))/length(which(patient.info$Status == 'NP'))*100`% of the NP cellularities missing.

3. With regards to p53 staining, non-progressors have no p53 positive tests, while progressors had `r p53Status[2,2]` overall.
`r pander(p53Status, justify='left', caption="P53 staining status for progressors vs non")`
Of these `r table(be$p53.Status)[['1']]` were positive in a BE sample, and `r table(lgd$p53.Status)[['1']]` were positive when LGD samples were included.

@@

  plots = list()
  for (sv in c('seg.vals','norm.seg.vals')) {

    x1 = patient.data[[pt]][[sv]][, patient.data[[pt]]$info$Samplename]
    x1 = x1[,patient.data[[pt]]$info$Samplename]
    colnames(x1) = paste(patient.data[[pt]]$info$Endoscopy.Year, '(', patient.data[[pt]]$info$Pathology, ')', sep='' )

    eigenvalues = pca$sdev^2
    screeplot(pca, main=paste(pt, df, 'scree'))

    gg = ggplot(df, aes(PC1, PC2)) + geom_line() +
        geom_label(aes(label=year),nudge_y=0.04) + 
        geom_point( aes(color=Pathology, size=Pathology)) +
        labs(title=paste(pt, sv))
  plots[[sv]] = gg
  }  
  do.call(grid.arrange, c(plots, list(ncol=2))
  return(gg)
}

plot.pca.norm<-function(pt) {

  x1 = patient.data[[pt]]$norm.seg.vals[, patient.data[[pt]]$info$Samplename]
  x1 = x1[,patient.data[[pt]]$info$Samplename]
  colnames(x1) = paste(patient.data[[pt]]$info$Endoscopy.Year, '(', patient.data[[pt]]$info$Pathology, ')', sep='' )

@@

    if (length(table(mod1dr$mixcomp)) > 1) {
      plot(mod1dr, what = "evalues")
      title(main=pt)
      #plot(mod1dr, what = "density")
    }
  })
}

@@

## Using the segmentation values
sim = sum.patient.data[,c('Status','years')]
rownames(sim) = sum.patient.data[,'Patient']
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  # if (ncol(x1) <= 15) {
  #   plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  # } else {
  #   plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  # }

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))

  tmp = patient.data[[pt]]$variable.regions[,-(1:5)]
  sim[pt, 'n.var'] = nrow(patient.data[[pt]]$variable.regions[,-(1:5)])

  if (nrow(tmp) <= 0) {
    sim[pt,c('gains', 'loss')] = c(0,0)
  } else {
    sim[pt,'gains'] = length(tmp[tmp>=1.1])
    sim[pt,'loss'] =  length(tmp[tmp<=0.9])
  }
}
sim$id = rownames(sim)

@@

## Number of variable regions isn't so good, but maybe...
sim %>%
  group_by(Status) %>%
  summarise( mean = mean(n.var), sd = sd(n.var), med = median(n.var))
t.test( subset(sim, Status == 'NP')$n.var, subset(sim, Status == 'P')$n.var )$p.value

@@

```{r xval_pt, echo=T, warning=F, message=T, fig.align='left', fig.height=4, eval=F}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

ggplot(melt(data.frame(loo_xval[,c(3)])), aes(x=variable, y=value, fill=variable)) +  
    geom_boxplot() + geom_jitter(shape=16, width=0.2) + geom_hline(yintercept=0.05, color='red', alpha=0.5) +
    stat_boxplot(geom ='errorbar') + labs(y='p-values', x='Iterations', title='Leave one (patient) out clustering') +
  scale_fill_manual(values=c('purple'), labels='clusters')

@@

```{r xval_smp, echo=F, warning=F, message=T, fig.align='left', fig.height=6, eval=F}

# Leave out samples?
looS_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:20) { 
  simX = sum.patient.data[,c('Status','years')]
  rownames(simX) = sum.patient.data[,'Patient']

@@

ggplot(melt(data.frame(looS_xval[,c(3)])), aes(x=variable, y=value, fill=variable)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
  geom_hline(yintercept=0.05, color='red', alpha=0.5) +
    stat_boxplot(geom ='errorbar') + labs(y='p-values', x='Iterations', title='Leave one (sample) out clustering') + 
    scale_fill_manual(values=c('#CC99FF'), labels='clusters')

@@

```{r nofinal, echo=F, message=F, warning=F, fig.height=6, fig.width=10}

gain.threshold = 1.1
loss.threshold = 0.9

simX = sum.patient.data[,c('Status','years')]
rownames(simX) = sum.patient.data[,'Patient']
# Leave out final sample on the progressors

for (pt in sum.patient.data$Patient) {
  print(pt)
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

  simX[pt, 'ns'] = net.similarity(ac)
  simX[pt, 'ss'] = sum.similarity(ac)
  simX[pt, 'clusters'] = length(clusters(ac))

  simX[pt, 'n.var'] = nrow(patient.data[[pt]]$variable.regions[,-(1:5)])
}

ttpc = t.test( subset(simX, Status == 'NP')$clusters, subset(simX, Status == 'P')$clusters )$p.value
ggplot(melt(data.frame(simX[c('Status','clusters')])), aes(x=Status, y=value, fill=Status)) +  
    geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
    stat_boxplot(geom ='errorbar') + labs(y='num. clusters', x='', title=paste('Leave out HGD sample(s) p-value', signif(ttpc, 3)))

t.test(subset(simX, Status == 'NP')$n.var, subset(simX, Status == 'P')$n.var)

@@

pander(subset(simX, Status == 'P' & clusters >= mean(clusters)), caption="These patients are more similar to Non-Progressors without their HGD sample(s)")

fit = glm(Status~clusters, simX, family=binomial(link='logit') )
summary(fit)

simX = simX[, c('Status', 'clusters', 'ns','ss','gain','loss')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  geom_vline(xintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

@@

pvals = as.data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('p.clusters','mean.clusters', 'sd.clusters'))))
status.clusters = list()
for (qv in seq(0,0.9,0.1)) {
  print(qv)

@@

    simPC[pt, 'ns'] = net.similarity(ac)
    #simPC[pt, 'ss'] = sum.similarity(ac)
    simPC[pt, 'clusters'] = length(clusters(ac))
  }

  # May be a difference
  #ttps = t.test( subset(simPC, Status == 'NP')$ss, subset(simPC, Status == 'P')$ss )$p.value
  ttpn = t.test( subset(simPC, Status == 'NP')$ns, subset(simPC, Status == 'P')$ns )$p.value
  ttpc = t.test( subset(simPC, Status == 'NP')$clusters, subset(simPC, Status == 'P')$clusters )$p.value

  pvals = rbind(pvals,cbind(ttpc, mean(simPC$clusters), sd(simPC$clusters)))

  by.clusters = simPC %>% group_by(Status) %>% summarise( n.mean.clusters=mean(clusters), n.sd.clusters=sd(clusters), se.clusters=sd(clusters)/sqrt(length(Status)) )

  status.clusters[[paste('q=',qv, sep='')]] = by.clusters
}
colnames(pvals) = c('p.clusters','mean.clusters', 'sd.clusters')
pvals$qv=seq(0,0.9,0.1)

clus = as.data.frame(do.call(rbind, lapply(status.clusters, function(x) c(x[['n.mean.clusters']], x[['n.sd.clusters']]  ))))
colnames(clus) = c('NP','P', 'NP.sd', 'P.sd')
clus$qv = seq(0,0.9,0.1)

m = melt(clus[,c(1:2,5)], measure.vars=c('NP','P'))
m[which(m$variable == 'NP'), 'sd'] = clus$NP.sd
m[which(m$variable == 'P'), 'sd'] = clus$P.sd

ggplot(m, aes(x=qv, y=value, col=variable)) + geom_point() +  
  geom_errorbar(aes(ymin=value-sd, ymax=value+sd), width=.01, alpha=0.3) +
  geom_hline(yintercept=5, alpha=0.3) + geom_hline(yintercept=15, alpha=0.3, col='red') + 
  labs(title='Clusters vs q parameters', x='q parameter', y='mean(clusters)')

ggplot(melt(pvals, measure.vars=c('p.clusters'), id.vars='qv'),aes(x=qv,y=value, color=variable)) + geom_point() +
  labs(title='P vs NP p-values per q parameters', y='pvalue NP vs P', x='q param')

@@

```{r model, echo=F, warning=F, message=F, fig.height=12, fig.width=20, eval=F}

t.test(subset(sim, Status == 'P')$clusters, subset(sim, Status == 'NP')$clusters)

t.test(subset(simX, Status == 'P')$clusters, subset(simX, Status == 'NP')$clusters)

## TODO  Need to consider segment lengths relative to chromosomal lengths

var.lengths = as.data.frame( do.call(rbind, lapply(patient.data, function(x){
  if (nrow(x[['variable.regions']]) > 0) {
    lengths = as.vector( (x[['variable.regions']]['end.pos'] - x[['variable.regions']]['start.pos'])[,1] )
    return(c(median(lengths), mean(lengths), sd(lengths), sd(lengths)/length(lengths)))
  }
  return(NA)
})) )
colnames(var.lengths) = c('median','mean','sd','se')

var.lengths = merge(sum.patient.data, var.lengths, by.x='Patient', by.y='row.names')

var.lengths
wilcox.test(subset(var.lengths, Status == 'P')$mean, subset(var.lengths, Status == 'NP')$mean)
